---
title: "Alzheimer data analysis"
output: html_document
date: "2026-09-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load libraries

```{r warning=FALSE echo = FALSE}

library(Seurat)
library(monocle3)
library(SeuratWrappers)
library(harmony)

library(UCell)
library(AUCell)


library(tidyr)
library(dplyr)

library(patchwork)
library(ggplot2)
library(RColorBrewer)


library(pcaMethods)
library(rstatix)
library(SingleR)


```







# Load data

<!-- Data description (Alzheimer) GSE175814 contains single-cell data from de cerveaux post-mortem de patients Alzheimer, avec plusieurs régions cérébrales. L'étude cherche notamment à caractériser les changements d'expression dans les neurones et cellules gliales --> 
<!-- Single cell RNA sequencing analysis of human Alzheimer’s disease brain samples reveal neuronal and glial specific cells differential expression -->

## Alzheimer Disease (AD)

### Patient 1


```{r}

AD1.data <- Read10X(data.dir = "C:/Users/kembi/Documents/Users/Patricia_Kalombo/BIOINFORMATICS/scRNAseq/DATA/GSE175814_RAW/AD_Patient_1/")

# Rename cells name
colnames(x = AD1.data) <- paste('AD1', colnames(x = AD1.data), sep = '_')


## Create Seurat object
AD1_Patient_1 <- CreateSeuratObject(counts = AD1.data, project = "AD1")

AD1_Patient_1

```



```{r}

# Lets examine a few genes in the first thirty cells
AD1.data[1:5, 1:30]

```


### Quality Control and Filter cells (remove cells of low quality): Remove cells with high number of mitochondrial genes & Remove cells with anormally high number of genes

```{r}

## Percentage of mito genes and store in "percent.mt" column
AD1_Patient_1[["percent.mt"]] <- PercentageFeatureSet(AD1_Patient_1, pattern = "^MT-")

head(AD1_Patient_1@meta.data, 5)

```




```{r warning=FALSE}

FeatureScatter(object = AD1_Patient_1, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(object = AD1_Patient_1, feature1 = "nFeature_RNA", feature2 = "percent.mt")
FeatureScatter(object = AD1_Patient_1, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```



```{r}

# Filter low quality cells
AD1_Patient_1 <- subset(x = AD1_Patient_1, subset = percent.mt < 10)

AD1_Patient_1

```



```{r}

## Check ater filtering
FeatureScatter(object = AD1_Patient_1, feature1 = "nFeature_RNA", feature2 = "percent.mt")

```



```{r}

AD1_Patient_1
#33538 features across 6201 samples within 1 assay 


```



### Patient 2

```{r}

AD2.data <- Read10X(data.dir = "C:/Users/kembi/Documents/Users/Patricia_Kalombo/BIOINFORMATICS/scRNAseq/DATA/GSE175814_RAW/AD_Patient_2/")

# Rename cells name
colnames(x = AD2.data) <- paste('AD2', colnames(x = AD2.data), sep = '_')


## Create Seurat object
AD2_Patient_2 <- CreateSeuratObject(counts = AD2.data, project = "AD2")

AD2_Patient_2

```



```{r}

# Lets examine a few genes in the first thirty cells
AD2.data[1:5, 1:30]

```


### Quality Control and Filter cells (remove cells of low quality): Remove cells with high number of mitochondrial genes & Remove cells with anormally high number of genes



```{r}

## Percentage of mito genes and store in "percent.mt" column
AD2_Patient_2[["percent.mt"]] <- PercentageFeatureSet(AD2_Patient_2, pattern = "^MT-")

head(AD2_Patient_2@meta.data, 5)

```




```{r warning=FALSE}

FeatureScatter(object = AD2_Patient_2, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(object = AD2_Patient_2, feature1 = "nFeature_RNA", feature2 = "percent.mt")
FeatureScatter(object = AD2_Patient_2, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```



```{r}

## Filter low quality cells
AD2_Patient_2 <- subset(x = AD2_Patient_2, subset = percent.mt < 20 & nFeature_RNA < 8000)
AD2_Patient_2

```



```{r}

## Check ater filtering
FeatureScatter(object = AD2_Patient_2, feature1 = "nFeature_RNA", feature2 = "percent.mt")

```



```{r}

AD2_Patient_2
#33538 features across 5187 samples within 1 assay 

```



## Healthy Donor (HD)

### Patient 1


```{r}

HD1.data <- Read10X(data.dir = "C:/Users/kembi/Documents/Users/Patricia_Kalombo/BIOINFORMATICS/scRNAseq/DATA/GSE175814_RAW/HD_Patient_1/")

# Rename cells name
colnames(x = HD1.data) <- paste('HD1', colnames(x = HD1.data), sep = '_')


## Create Seurat object
HD1_Patient_1 <- CreateSeuratObject(counts = HD1.data, project = "HD1")

HD1_Patient_1

```




```{r}

# Lets examine a few genes in the first thirty cells
HD1.data[1:5, 1:30]

```




### Quality Control and Filter cells (remove cells of low quality): Remove cells with high number of mitochondrial genes & Remove cells with anormally high number of genes

```{r}

## Percentage of mito genes and store in "percent.mt" column
HD1_Patient_1[["percent.mt"]] <- PercentageFeatureSet(HD1_Patient_1, pattern = "^MT-")

head(HD1_Patient_1@meta.data, 5)

```





```{r warning=FALSE}

FeatureScatter(object = HD1_Patient_1, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(object = HD1_Patient_1, feature1 = "nFeature_RNA", feature2 = "percent.mt")
FeatureScatter(object = HD1_Patient_1, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```




```{r}

## Filter low quality cells

HD1_Patient_1 <- subset(x = HD1_Patient_1, subset = percent.mt < 20 & nFeature_RNA < 8000)

HD1_Patient_1

```




```{r}

## Check ater filtering

FeatureScatter(object = HD1_Patient_1, feature1 = "nFeature_RNA", feature2 = "percent.mt")

```




```{r}

HD1_Patient_1
#33538 features across 5971 samples within 1 assay 

```

### Patient 2


```{r}

HD2.data <- Read10X(data.dir = "C:/Users/kembi/Documents/Users/Patricia_Kalombo/BIOINFORMATICS/scRNAseq/DATA/GSE175814_RAW/HD_Patient_2/")

# Rename cells name
colnames(x = HD2.data) <- paste('HD2', colnames(x = HD2.data), sep = '_')


## Create Seurat object
HD2_Patient_2 <- CreateSeuratObject(counts = HD2.data, project = "HD2")

HD2_Patient_2

```




```{r}

# Lets examine a few genes in the first thirty cells
HD2.data[1:5, 1:30]

```




### Quality Control and Filter cells (remove cells of low quality): Remove cells with high number of mitochondrial genes & Remove cells with anormally high number of genes

```{r}

## Percentage of mito genes and store in "percent.mt" column
HD2_Patient_2[["percent.mt"]] <- PercentageFeatureSet(HD2_Patient_2, pattern = "^MT-")

head(HD2_Patient_2@meta.data, 5)

```





```{r warning=FALSE}

FeatureScatter(object = HD2_Patient_2, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(object = HD2_Patient_2, feature1 = "nFeature_RNA", feature2 = "percent.mt")
FeatureScatter(object = HD2_Patient_2, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```




```{r}

## Filter low quality cells
HD2_Patient_2 <- subset(x = HD2_Patient_2, subset = percent.mt < 10 & nFeature_RNA < 10800)

HD2_Patient_2

```




```{r}

## Check ater filtering
FeatureScatter(object = HD2_Patient_2, feature1 = "nFeature_RNA", feature2 = "percent.mt")

```



```{r}

HD2_Patient_2
#33538 features across 10924 samples within 1 assay 

```




# Merge datasets (seurat object)

```{r}

merged_seurat_data <- merge(AD1_Patient_1,
                            y = list(AD2_Patient_2, HD1_Patient_1, HD2_Patient_2),
                            project = "Merged_AD_1_2_HD_1_2")

# merged_seurat_data <- merge(AD1_Patient_1, y = HD1_Patient_1, 
#                             project = "Merged_AD_1_HD_1")
merged_seurat_data

```


```{r}

merged_seurat_data$Samples <- rownames(merged_seurat_data@meta.data)

merged_seurat_data@meta.data <- separate(merged_seurat_data@meta.data, col = 'Samples', into = c('Patient', 'Barcode'), sep = '_')

merged_seurat_data@meta.data$Patient_Condition <- "Healthy"


merged_seurat_data@meta.data$Patient_Condition[merged_seurat_data@meta.data$Patient %in% c("AD1", "AD2")] <- "Alzheimer"

# View(merged_seurat_data@meta.data)

```



## Check quality control on merged dataset

```{r}

FeatureScatter(object = merged_seurat_data, feature1 = "nFeature_RNA", feature2 = "percent.mt")

```



```{r}

merged_seurat_data
#33538 features across 28283 samples within 1 assay

```





# Data integration

## CCA methods

### Preprocess data without integration


```{r warning=FALSE}

## Normalize and Transform data & Find fatures highly variables
merged_seurat_data_integration <- SCTransform(object = merged_seurat_data)

## PCA
merged_seurat_data_integration <- RunPCA(merged_seurat_data_integration)
## Help to choose number of components
ElbowPlot(merged_seurat_data_integration)


merged_seurat_data_integration <- FindNeighbors(merged_seurat_data_integration, dims = 1:20, reduction = "pca", verbose = FALSE)
merged_seurat_filtered <- FindClusters(merged_seurat_data_integration)

merged_seurat_data_integration <- RunUMAP(merged_seurat_data_integration, dims = 1:20, reduction.name = "UMAP_unintegrated")


DimPlot(merged_seurat_data_integration, reduction = "UMAP_unintegrated", label = TRUE)

DimPlot(merged_seurat_data_integration, reduction = "UMAP_unintegrated", group.by = "Patient")


```

### Perform integration with CCA methods


```{r}

merged_seurat_data_integration <- IntegrateLayers(object = merged_seurat_data_integration, method = CCAIntegration, normalization.method = "SCT", verbose = F)

merged_seurat_data_integration <- FindNeighbors(merged_seurat_data_integration, reduction = "integrated.dr", dims = 1:20)

merged_seurat_data_integration <- FindClusters(merged_seurat_data_integration, resolution = 0.05)


merged_seurat_data_integration <- RunUMAP(merged_seurat_data_integration, dims = 1:20, reduction = "integrated.dr", reduction.name = "UMAP_integrated")


## Vizualisation
DimPlot(merged_seurat_data_integration, reduction = "UMAP_integrated", group.by = c("Patient")) + ggtitle("AD_HD")

DimPlot(merged_seurat_data_integration, reduction = "UMAP_integrated", label = TRUE)


DimPlot(merged_seurat_data_integration, reduction = "UMAP_integrated", group.by = c("Patient", "seurat_clusters"))



DimPlot(merged_seurat_data_integration, reduction = "UMAP_integrated", split.by = "Patient", label = TRUE) + ggtitle("Alzheimer") + theme(plot.title = element_text(hjust = 0.5))  + labs(x ="UMAP_1", y = "UMAP_2")

# p_umap <- DimPlot(merged_seurat_integration, reduction = "UMAP_integrated", split.by = "Patient", label = TRUE) + ggtitle("Alzheimer") + theme(plot.title = element_text(hjust = 0.5))  + labs(x ="UMAP_1", y = "UMAP_2")

# ggsave(p_umap,file = "C:/Users/kembi/Documents/Users/Patricia_Kalombo/CRCT/TUTORIAL/GSEA/Sabatier2023/RESULTS/scRNAseq/PDX_TUH110/UMAP_Alzheimer.pdf", device = "pdf", width = 7, height = 5)



```


```{r}


## Vizualisation


# p_umap <- DimPlot(merged_seurat_integration, reduction = "UMAP_integrated", label = TRUE) + ggtitle("Alzheimer Disease") + theme(plot.title = element_text(hjust = 0.5))  + labs(x ="UMAP_1", y = "UMAP_2")

# p_umap <- DimPlot(merged_seurat_integration, reduction = "UMAP_integrated", label = TRUE) + ggtitle("") + theme(plot.title = element_text(hjust = 0.5))  + labs(x ="UMAP_1", y = "UMAP_2")

p_umap <- DimPlot(merged_seurat_integration, reduction = "UMAP_integrated", label = TRUE) + NoLegend() + NoAxes() #ggtitle("") + theme(plot.title = element_text(hjust = 0.5))  + labs(x ="UMAP_1", y = "UMAP_2")
# 
# p_umap <- DimPlot(merged_seurat_integration, reduction = "UMAP_integrated", split.by = "Patient", label = TRUE) + ggtitle("PDX_TUH110_GILT_D7") + theme(plot.title = element_text(hjust = 0.5))  + labs(x ="UMAP_1", y = "UMAP_2")

ggsave(p_umap,file = "C:/Users/kembi/Documents/Users/Patricia_Kalombo/BIOINFORMATICS/scRNAseq/RESULTS/alzheimer_disease_2AD_2HD.jpeg", width = 7, height = 5, dpi = 300)

# ggsave("mon_plot.jpeg", plot = p, width = 8, height = 6)

```



### Save seurat object

```{r}

## save results of integration with CCA methods
# saveRDS(merged_seurat_integration, file = "C:/Users/kembi/Documents/Users/Patricia_Kalombo//merged_seurat_integration_005.rds")

```



### Read RDS seurat object

```{r}

# merged_seurat_integration <- readRDS("C:/Users/kembi/Documents/Users/Patricia_Kalombo/merged_seurat_integration_005.rds")

```




# Cell type annotation using genes markers

```{r}

##List of genes markers for annotations
markers_neuro <- list(
  # Microglia
  Microglia = c("P2RY12", "TMEM119","CX3CR1", "AIF1", "C1QA",
                "C1QB", "C1QC", "SALL1", "TYROBP", "LST1"),

  # Microglia activated / DAM
  Microglia_DAM = c("APOE", "TREM2", "GPNMB", "SPP1", "CST7",
                    "LPL", "LGALS3", "CLEC7A"),

  # Astrocytes
  Astrocytes = c("GFAP", "AQP4", "ALDH1L1", "SLC1A3",
                 "S100B",  "SOX9", "FABP7"),

  # Reactive astrocytes
  Astrocytes_reactive = c("GFAP", "VIM", "C3",
                          "SERPINA3", "LCN2", "CLU"),

  # Oligodendrocytes
  Oligodendrocytes = c("MBP", "PLP1",  "MOG",  "MAG",
                       "MOBP", "CLDN11",  "OPALIN"),

  # OPC
  OPC = c("PDGFRA", "CSPG4", "OLIG1", "OLIG2","SOX10" ),

  # Excitatory neurons
  Excitatory_neurons = c("SLC17A7", "SLC17A6", "CAMK2A", "SATB2","TBR1"),

  # Inhibitory neurons
  Inhibitory_neurons = c("GAD1", "GAD2", "SLC32A1","DLX1", "DLX2"),

  # PVALB interneurons
  PVALB = c("PVALB", "GAD1", "GAD2" ),

  # SST interneurons
  SST = c("SST", "GAD1", "GAD2"),

  # VIP interneurons
  VIP = c("VIP", "GAD1","GAD2"),

  # Endothelial
  Endothelial = c("CLDN5", "PECAM1", "VWF", "KDR", "EMCN", "RAMP2"),

  # Pericytes
  Pericytes = c("PDGFRB", "RGS5", "CSPG4", "MCAM", "DES"),

  # Ependymal
  Ependymal = c( "FOXJ1", "PIFO", "TPPP3", "RSPH1" )
)


## Keep only genes that are in normalized matrix
markers_neuro <- lapply(markers_neuro, function(x) {
  intersect(x, rownames(merged_seurat_data_integration))
})



```



```{r}


markers_neuro <- list(
  # Microglia
  Microglia = c("P2RY12", "TMEM119","CX3CR1", "AIF1", "C1QA",
                "C1QB", "C1QC", "SALL1", "TYROBP", "LST1"),

  # # Microglia activated / DAM
  # Microglia_DAM = c("APOE", "TREM2", "GPNMB", "SPP1", "CST7",
  #                   "LPL", "LGALS3", "CLEC7A"),

  # Astrocytes
  Astrocytes = c("GFAP", "AQP4", "ALDH1L1", "SLC1A3",
                 "S100B",  "SOX9", "FABP7"),

  # # Reactive astrocytes
  # Astrocytes_reactive = c("GFAP", "VIM", "C3",
  #                         "SERPINA3", "LCN2", "CLU"),

  # Oligodendrocytes
  Oligodendrocytes = c("MBP", "PLP1",  "MOG",  "MAG",
                       "MOBP", "CLDN11",  "OPALIN"),

  # OPC
  OPC = c("PDGFRA", "CSPG4", "OLIG1", "OLIG2","SOX10" ),

  # Excitatory neurons
  Excitatory_neurons = c("SLC17A7", "SLC17A6", "CAMK2A", "SATB2","TBR1"),

  # Inhibitory neurons
  Inhibitory_neurons = c("GAD1", "GAD2", "SLC32A1","DLX1", "DLX2"),

  # PVALB interneurons
  # PVALB = c("PVALB", "GAD1", "GAD2" ),

  # SST interneurons
  # SST = c("SST", "GAD1", "GAD2"),

  # VIP interneurons
  # VIP = c("VIP", "GAD1","GAD2"),
  
  PVALB_SST_VIP = c("PVALB", "GAD1", "GAD2", "SST", "VIP" ),

  # Endothelial
  Endothelial = c("CLDN5", "PECAM1", "VWF", "KDR", "EMCN", "RAMP2"),

  # Pericytes
  Pericytes = c("PDGFRB", "RGS5", "CSPG4", "MCAM", "DES"),

  # Ependymal
  Ependymal = c( "FOXJ1", "PIFO", "TPPP3", "RSPH1" )
)


## Keep only genes that are in normalized matrix
markers_neuro <- lapply(markers_neuro, function(x) {
  intersect(x, rownames(merged_seurat_data_integration))
})



```



## Compute score for each populations with AddModuleScore function

```{r}

merged_seurat_data_integration <- AddModuleScore(object = merged_seurat_data_integration,
                                                 features = markers_neuro,
                                                 name =  paste0(names(markers_neuro), "_score"),
                                                 ctrl = 100,  nbin = 24, seed = 123)



```


```{r}

DotPlot(
  merged_seurat_data_integration,
  features = c(
    "Microglia_score1", "Microglia_DAM_score2", 
    "Astrocytes_score3", "Astrocytes_reactive_score4",
    "Oligodendrocytes_score5", "OPC_score6", 
     "Excitatory_neurons_score7",
    "Inhibitory_neurons_score8"
  ),
  group.by = "seurat_clusters"
) +
  RotatedAxis() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )


```



```{r}

score_cols <- grep("_score", colnames(merged_seurat_data_integration@meta.data), value = TRUE)


# Do cell type annotation based on the high score for each cell
merged_seurat_data_integration$Celltype <- apply(
  merged_seurat_data_integration@meta.data[, score_cols],
  1,
  function(x) names(markers_neuro)[which.max(x)]
)


# UMAP by cell type annoted
DimPlot(merged_seurat_data_integration, group.by = "Celltype", label = TRUE) + 
  ggtitle("Cell type annotation using geneset markers") + NoLegend()

DimPlot(merged_seurat_data_integration, group.by = "Celltype", label = TRUE) + 
  ggtitle("Cell type annotation using geneset markers")


```




### Proportion of cells by condition for each cluster

```{r}

donnees <- merged_seurat_data_integration@meta.data[, c("Patient", "Celltype", "seurat_clusters")]

donnees <- donnees %>%
  group_by(seurat_clusters, Celltype) %>%  ## , Patient
  summarise(Proportion = n())

colnames(donnees) <- sub("Celltype","Cells_Pops", colnames(donnees))


pClusters <- ggplot(donnees, aes( x=seurat_clusters, y=Proportion, fill=Cells_Pops)) +
  geom_bar(position="fill", stat="identity") +
  theme_classic() + labs(x = "Clusters", y = "Fraction of cells in each cluster") +
  ggtitle("Cluster composition") +
  theme(plot.title = element_text(hjust = 0.5))


# ggsave(pClusters,file = "C:/Users/kembi/Documents/Users/Patricia_Kalombo/BIOINFORMATICS/scRNAseq/RESULTS/alzheimer_disease_2AD_2HD_barplot.jpeg", width = 7, height = 5, dpi = 300)


```



```{r}

# p_umap_cluster <- p_umap | pClusters
# 
# ggsave(p_umap_cluster,file = "C:/Users/kembi/Documents/Users/Patricia_Kalombo/BIOINFORMATICS/scRNAseq/RESULTS/alzheimer_disease_2AD_2HD_umap_barplot.jpeg", width = 9, height = 5, dpi = 300)

```



##  Gene Set Enrichment Analysis with CCA results (enrichment signature score)


```{r}

## INFLAMMATORY_SIGNATURES

signature_genes <- list(INFLAMMATORY_SIGNATURES = 
                          c("ABCA1","ABI1","ACVR1B","ACVR2A","ADM","ADORA2B","ADRM1","AHR",
                        "APLNR","AQP9","ATP2A2","ATP2B1","ATP2C1","AXL","BDKRB1","BEST1",
                        "BST2","BTG2","C3AR1","C5AR1","CALCRL","CCL17","CCL2","CCL20",
                        "CCL22","CCL24","CCL5","CCL7","CCR7","CCRL2","CD14","CD40","CD48",
                        "CD55","CD69","CD70","CD82","CDKN1A","CHST2","CLEC5A","CMKLR1",
                        "CSF1","CSF3","CSF3R","CX3CL1","CXCL10","CXCL11","CXCL6","CXCL9",
                        "CXCR6","CYBB","DCBLD2","EBI3","EDN1","EIF2AK2","EMP3","ADGRE1",
                        "EREG","F3","FFAR2","FPR1","FZD5","GABBR1","GCH1","GNA15","GNAI3",
                        "GP1BA","GPC3","GPR132","GPR183","HAS2","HBEGF","HIF1A","HPN",
                        "HRH1","ICAM1","ICAM4","ICOSLG","IFITM1","IFNAR1","IFNGR2","IL10",
                        "IL10RA","IL12B","IL15","IL15RA","IL18","IL18R1","IL18RAP","IL1A",
                        "IL1B","IL1R1","IL2RB","IL4R","IL6","IL7R","CXCL8","INHBA","IRAK2",
                        "IRF1","IRF7","ITGA5","ITGB3","ITGB8","KCNA3","KCNJ2","KCNMB2","KIF1B",
                        "KLF6","LAMP3","LCK","LCP2","LDLR","LIF","LPAR1","LTA","LY6E","LYN",
                        "MARCO","MEFV","MEP1A","MET","MMP14","MSR1","MXD1","MYC","NAMPT",
                        "NDP","NFKB1","NFKBIA","NLRP3","NMI","NMUR1","NOD2","NPFFR2","OLR1",
                        "OPRK1","OSM","OSMR","P2RX4","P2RX7","P2RY2","PCDH7","PDE4B","PDPN",
                        "PIK3R5","PLAUR","PROK2","PSEN1","PTAFR","PTGER2","PTGER4","PTGIR",
                        "PTPRE","PVR","RAF1","RASGRP1","RELA","RGS1","RGS16","RHOG","RIPK2",
                        "RNF144B","ROS1","RTP4","SCARF1","SCN1B","SELE","SELL","SELENOS",
                        "SEMA4D","SERPINE1","SGMS2","SLAMF1","SLC11A2","SLC1A2","SLC28A2",
                        "SLC31A1","SLC31A2","SLC4A4","SLC7A1","SLC7A2","SPHK1","SRI","STAB1",
                        "TACR1","TACR3","TAPBP","TIMP1","TLR1","TLR2","TLR3","TNFAIP6",
                        "TNFRSF1B","TNFRSF9","TNFSF10","TNFSF15","TNFSF9","TPBG","VIP"))


## Inflammatory  geneset signatures
signature_genes <- list(c(signature_genes$INFLAMMATORY_SIGNATURES))

## Keep only genes expressed in data
signature_genes <- lapply(signature_genes, function(x) {
  intersect(x, rownames(merged_seurat_data_integration))
})

signature_genes[which(sapply(signature_genes, length) == 0)] <- NULL


merged_seurat_data_integration <- AddModuleScore(merged_seurat_data_integration, 
                                                 features = signature_genes, 
                                                 name = "INFLAMMATORY_SIGNATURES", 
                                                 nbin =  20)



# View(merged_seurat_data_integration@meta.data)



```








## Vizualisation

### UMAP

```{r}


## Visualize score on UMAP

## INFLAMMATORY_SIGNATURES
FeaturePlot(merged_seurat_data_integration, features = "INFLAMMATORY_SIGNATURES1", label = TRUE, reduction = "UMAP_integrated",raster= FALSE,label.size = 6, repel = TRUE, cols = c("grey",   "red")) #+ scale_colour_gradientn(colours= rev(brewer.pal(n=11, name = "Spectral")))

FeaturePlot(merged_seurat_data_integration, features = "INFLAMMATORY_SIGNATURES1", split.by = "Patient", label = TRUE, reduction = "UMAP_integrated", raster= FALSE,label.size = 6, repel = TRUE) #+ scale_colour_gradientn(colours= rev(brewer.pal(n=11, name = "Spectral")))



```



#### Which populations expressed interests Genes 


```{r}

FeaturePlot(merged_seurat_data_integration, features = c("APBB1IP", "ST18","SV2B"), label = TRUE, reduction = "UMAP_integrated", raster= FALSE,label.size = 5, repel = TRUE, cols = c("grey",   "red"))

FeaturePlot(merged_seurat_data_integration, features = c("GRIK3", "PDGFRA"), label = TRUE, reduction = "UMAP_integrated", raster= FALSE,label.size = 5, repel = TRUE, cols = c("grey",   "red"))


###, min.cutoff = -2, max.cutoff = 2 split.by = "Patient", 

```




# MONOCLE3 WORKFLOW


```{r}

# 1 Convert  seurat object to cell_data_set object for monocle3

cds <- as.cell_data_set(merged_seurat_data_integration)
cds


```


```{r}

# to get cell metadata
colData(cds)
# to gene metdata
fData(cds)
rownames(fData(cds))[1:10]

# since it misses the gene_short_name column, let's add it
fData(cds)$gene_short_name <- rownames(fData(cds))

# to get counts
counts(cds)


```


## Cluster cells (using clustering info from seurat's UMAP)


```{r}


# 2. Cluster cells (using clustering info from seurat's UMAP)
# let's use the clustering information have

# 1. Assign paritions
reacreate.partition <- c(rep(1,length(cds@colData@rownames)))
names(reacreate.partition) <- cds@colData@rownames
reacreate.partition <- as.factor(reacreate.partition)

cds@clusters$UMAP$partitions <- reacreate.partition


# 2. Assign the cluster info 
list_cluster <- merged_seurat_data_integration@active.ident
cds@clusters$UMAP$clusters <- list_cluster


# Assign UMAP coordinate - cell embeddings
cds@int_colData@listData$reducedDims$UMAP <- merged_seurat_data_integration@reductions$UMAP_integrated@cell.embeddings




```





```{r}


# Plot

cluster.before.trajectory <- plot_cells(cds,
                                        color_cells_by = 'cluster',
                                        label_groups_by_cluster = FALSE,
                                        group_label_size = 5) +
  theme(legend.position = "right")

cluster.names <- plot_cells(cds,
           color_cells_by = "Celltype",
           label_groups_by_cluster = FALSE,
           group_label_size = 5) +
  # scale_color_manual(values = c('red', 'blue', 'green', 'maroon', 'yellow', 'grey', 'cyan')) +
  theme(legend.position = "none")

cluster.before.trajectory | cluster.names





```


### Trajectory analysis


```{r}


# 3. Learn trajectory graph
cds <- learn_graph(cds, use_partition = FALSE)

plot_cells(cds,
           color_cells_by = 'Celltype',
           label_groups_by_cluster = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE,
           group_label_size = 5)



```


### Psuedotime: 4. Order the cells in pseudotime

```{r}


# 4. Order the cells in pseudotime

cds <- order_cells(cds, reduction_method = 'UMAP', 
                    root_cells = colnames(cds[,clusters(cds) == 1]))

plot_cells(cds,
           color_cells_by = 'pseudotime',
           label_groups_by_cluster = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE)






```

```{r}

# cells ordered by monocle3 pseudotime

pseudotime(cds)
cds$monocle3_pseudotime <- pseudotime(cds)
data.pseudo <- as.data.frame(colData(cds))



ggplot(data.pseudo, aes(monocle3_pseudotime, reorder(Celltype, monocle3_pseudotime, median), fill = Celltype)) +
  geom_boxplot() 



```




## Finding differentially expressed features for all cluster

```{r}

merged_seurat_data_integration_markers <- PrepSCTFindMarkers(merged_seurat_data_integration)


```



```{r}



library(future)

plan("multicore", workers = 2)  # ou "multisession" sur Windows
options(future.globals.maxSize = 3 * 1024^3)  # augmente la limite to 10 Go



genesDEGs_clusters <- FindAllMarkers(merged_seurat_data_integration_markers,verbose = FALSE, only.pos = FALSE, logfc.threshold = 0) #, test.use = "DESeq2")


genesDEGs_clusters <- data.frame(Genes =  rownames(genesDEGs_clusters),
                           genesDEGs_clusters)


genesDEGs_clusters %>%
  group_by(cluster) %>%
  arrange()

```


```{r}

library(dplyr)

top5_genes <- genesDEGs_clusters %>%
  filter(avg_log2FC > 0) %>%              # uniquement les gènes upregulated
  filter(p_val_adj < 0.05) %>%           # optionnel : significatifs
  group_by(cluster) %>%
  slice_max(
    order_by = avg_log2FC,
    n = 5,
    with_ties = FALSE
  ) %>%
  ungroup()

top5_genes



```





```{r}

genes <- unique(top5_genes$gene)

genes <- top5_genes %>%
  arrange(cluster, desc(avg_log2FC)) %>%
  pull(gene) %>%
  unique()

## DotPlot
DotPlot(
  object = merged_seurat_data_integration,
  features = genes) +
  RotatedAxis()


```




